---
title: "Snap bean field locations"
subtitle: "with weather stations"
author: "Denis Shah"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r libraries}
library(tidyverse)
library(maps)
# library(sf)
# library(viridis)
# library(terra)
```


# Data preparation
```{r data-preparation}
# The observational matrix:
load(here::here("Data", "Survey.RData"))  # df

# Data frame of the fields and their locations:
surv <-
  df %>% 
  # Filter out the PA fields (Potter county):
  dplyr::filter(! county == "Potter") %>% 
  dplyr::select(subject, latitude, longitude, sampling.date, wm) %>% 
  dplyr::filter(!is.na(latitude), !is.na(longitude)) %>% 
  dplyr::arrange(subject, sampling.date) %>% 
  dplyr::group_by(subject) %>%
  # The last sampling date for each field:
  dplyr::slice_max(sampling.date, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>% 
  dplyr::filter(!is.na(wm)) %>% 
  dplyr::mutate(wm = ifelse(wm > 0, 1, 0)) %>% 
  dplyr::select(subject, latitude, longitude)


# Data frame of weather stations in NY. It is called st and the location columns are LAT and LON
load(here::here("Maps", "NYStations.RData"))  # st

# No outliers:
st %>% 
  dplyr::filter(LAT > 45 | LAT < 39)

# st <-
#   st %>% 
#   dplyr::filter(!NAME == "NIAGARA FALLS")

# The counties in which the surveyed fields were located
counties <-
  df %>% 
  # Filter out the PA fields (Potter county):
  dplyr::filter(! county == "Potter") %>% 
  dplyr::select(county) %>% 
  dplyr::distinct() %>% 
  dplyr::pull() %>% 
  as.character()

# Counties in lowercase:
counties_lc <- tolower(counties)
```


# Example using the `maps` package
```{r maps-package-example, eval=FALSE}
# This produces a simple map of the state. I used this as a check on the ranges of lat and lon, which were then used to check the st data frame for outliers
maps::map('county', region = 'new york', col = "#5E610B")
map.axes(cex.axis = 0.8) 
title(xlab = "Longitude", ylab = "Latitude", cex.lab = 0.8)
```


# Example using ggplot2
Use the function `map_data` from `ggplot2`, which converts the map to a data frame. 
The most important variable we need to pass to map_data is the name of map provided by the `maps` package.

```{r ggplot2-example, eval=FALSE}
ny_map_data <- map_data('county', region = 'new york')

ggplot() + 
  geom_polygon(data = ny_map_data, aes(x = long, y = lat, group = group), color = "black", fill = "#fafced") +
  labs(x = "longitude", y = "latitude") +
  coord_fixed(1.3) +
  theme(panel.background = element_blank())
```


# Expand to color only the counties in which we had samples
```{r ggplot2-example-expanded, eval=FALSE}
ggplot() + 
  geom_polygon(data = ny_map_data, 
               aes(x = long, y = lat, group = group), 
               color = "black", 
               fill = "white") +
  geom_polygon(data = ny_map_data %>% dplyr::filter(subregion %in% counties_lc), 
               aes(x = long, y = lat, group = group), 
               color = "black", 
               fill = "#fafced") +
  # Fix for the point colors
  geom_point(data = surv,
             aes(longitude, latitude),
             color = "blue3",  # Direct color assignment
             size = 1) +
  geom_point(data = st,
             aes(LON, LAT),
             color = "red3",  # Direct color assignment
             size = 1,
             pch = 8) +
  labs(x = "longitude", y = "latitude") +
  coord_fixed(1.3, 
              xlim = c(-80,-76.7),
              ylim = c(41.8, 43.5)) +
  theme(panel.background = element_blank())
```


# Using ggplot, sf and tigris shapefile
This is the best approach I think. Fields and stations do not appear in the Lakes!
We are able to set the crs.

```{r finalized-map-option}
library(tidyverse)
library(sf)
library(tigris)

# Download New York state counties
ny_counties <- counties(state = "NY", cb = TRUE) %>%
  st_transform(crs = 4326)

# Create a vector of counties to highlight
highlight_counties <- counties

# Add a column for fill color
ny_counties$fill_color <- ifelse(ny_counties$NAME %in% highlight_counties, "#fafced", NA)

# Convert surv dataframe to sf object
surv_sf <- st_as_sf(surv, coords = c("longitude", "latitude"), crs = 4326)

# Convert st dataframe to sf object
st_sf <- st_as_sf(st, coords = c("LON", "LAT"), crs = 4326)

# Create the map
ggplot() +
  geom_sf(data = ny_counties, aes(fill = fill_color), color = "black") +
  geom_sf(data = surv_sf, color = "blue", size = 1) +
  geom_sf(data = st_sf, color = "red3", size = 2, pch = 8) +
  scale_fill_identity() +
  theme(panel.background = element_blank()) +
  coord_sf(xlim = c(-80, -76.7), ylim = c(41.8, 43.5))
  
```










