---
title: "Fuse the white mold datasets"
subtitle: "Observational, agronomic, soils and weather variables"
author: "Denis Shah"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_caption: yes
    highlight: tango
    number_sections: true
    theme: cerulean
    toc: yes
    toc_depth: 4
    toc_float: true
    editor_options: 
      chunk_output_type: console
---

# OBJECTIVE

Fuse the different sources of variables linked to the field observations: - the survey data (subject and wm response variable) - agronomic (drainage, hydrol, year, cd, harv.optim) - soils (ph, om, log_sand_clay, log_silt_clay) - canopy closure at 35 dap - total rain from planting to 35 dap (pre-bloom) - total rain from 36-50 dap (bloom through early pod fill) - vars identified from FDA: - t2m_mean: start = 0, end = 4 - sm: start = -4, end = 3 - sm: start = 5, end = 15 - sm: start = 17, end = 24 - sm: start = 40, end = 49 - stsm: start = 35, end = 44

```{r setup, include=FALSE, eval=TRUE}
options(digits = 3)
require(knitr)
## options
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, cache = FALSE, warning = FALSE, message = FALSE)
```

```{r Libraries, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```

# the survey data (subject and wm response variable)

```{r}
# The observational (survey) matrix:
load(here::here("Data", "Survey.RData"))  # df

surv <-
  df %>% 
  # Filter out the PA fields (Potter county):
  dplyr::filter(! county == "Potter") %>% 
  dplyr::select(subject, latitude, longitude, sampling.date, wm) %>% 
  dplyr::filter(!is.na(latitude), !is.na(longitude)) %>% 
  dplyr::arrange(subject, sampling.date) %>% 
  dplyr::group_by(subject) %>%
  # The last sampling date for each field:
  dplyr::slice_max(sampling.date, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>% 
  dplyr::filter(!is.na(wm)) %>% 
  dplyr::mutate(wm = ifelse(wm > 0, 1, 0)) %>% 
  dplyr::select(subject, wm)
```

# agronomic (drainage, hydrol, cd, harv.optim)

```{r}
agron <-
  df %>% 
  dplyr::filter(! county == "Potter") %>% 
  # Filter out the missing location data:
  dplyr::filter(!is.na(latitude), !is.na(longitude)) %>% 
  dplyr::group_by(subject) %>%
  # The last sampling date for each field:
  dplyr::slice_max(sampling.date, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>% 
  # Collapse drainage.class into two groups:
  dplyr::mutate(drainage = 
                  forcats::fct_collapse(drainage.class,
                                        `Poorly_Drained` = c("Somewhat poorly drained", "Poorly drained", "Very poorly drained"),
                                        `Well_Drained` = c("Somewhat excessively drained", "Well drained", "Moderately well drained"))) %>%
  # Collapse the dual categories of hydro.group into group D (natural condition):
  dplyr::mutate(hydrol = 
                  forcats::fct_collapse(hydro.group,
                                        `A` = "A",
                                        `B` = "B",
                                        `C` = "C",
                                        `D` = c("D", "A/D", "B/D", "C/D"))) %>%
  # If dap is >60, then consider the field beyond the optimal harvest time (60 dap): Create a binary variable to represent this:
  dplyr::mutate(harv.optim = ifelse(dap <= 60, 0, 1)) %>% 
  dplyr::mutate(harv.optim = factor(harv.optim, levels = c(0, 1), labels = c("Yes", "No"))) %>% 
  # Selecting vars with no missing values, and which don't have a lot of small obs in categories:
  dplyr::select(subject, drainage, hydrol, cd, harv.optim) %>% 
  # Removal of duplicated rows:
  dplyr::distinct()
```

# soils (ph, om, log_sand_clay, log_silt_clay)

```{r}
soils <- 
  read.csv(here::here("Data", "extracted_soil_data.csv")) %>% 
  dplyr::select(-longitude, -latitude) %>% 
  # Rescale sand, silt, clay so that they add to 100 while respecting the proportionality among them
  dplyr::mutate(scaling_factor = 100/(clay+sand+silt)) %>% 
  dplyr::mutate(across(c(clay, sand, silt), ~ .x*scaling_factor)) %>% 
  dplyr::select(subject, ph, om, sand, silt, clay) %>% 
  # We want log ratios for sand, silt, clay as they are compositional.
  # Will use clay as the reference:
  dplyr::mutate(log_sand_clay = log(sand/clay)) %>% 
  dplyr::mutate(log_silt_clay = log(silt/clay)) %>% 
  dplyr::select(subject, ph, om, log_sand_clay, log_silt_clay)
```

# canopy closure at 35 dap

```{r}
load(here::here("CanopyClosure", "cc.df.35dap.RData")) # cc.df.35dap
```

# Rain

-   total rain from planting to 35 dap (pre-bloom)
-   total rain from 36-50 dap (bloom through early pod fill)

```{r}
load(here::here("Openmeteo", "rain_vars.RData")) # rain_vars
```

# Other environmental variables

-   vars identified from FDA:
    -   t2m_mean: start = 0, end = 4
    -   sm: start = -4, end = 3
    -   sm: start = 5, end = 15
    -   sm: start = 17, end = 24
    -   sm: start = 40, end = 49
    -   stsm: start = 35, end = 44

```{r}
load(here::here("FunctionalDataAnalysis", "FunctiononScalar", "WeatherVars.RData"))  # weather.vars
```

# Join em up!

```{r}
X <- purrr::reduce(list(surv, agron, soils, cc.df.35dap, rain_vars, weather.vars), dplyr::left_join, by = "subject") %>% 
  # Rename vars to avoid things like a minus sign in a var name (avoid possible problems later on), more descriptive names. dbp = days before planting, dap = days after planting
   dplyr::rename(t2m_mean_to_4dap = `t2m_mean_0_4`) %>%
   dplyr::rename(sm_4dbp_to_3dap = `sm_-4_3`) %>% 
   dplyr::rename(sm_5dap_to_15dap = `sm_5_15`) %>% 
   dplyr::rename(sm_17dap_to_24dap = `sm_17_24`) %>% 
   dplyr::rename(sm_40dap_to_49dap = `sm_40_49`) %>% 
   dplyr::rename(stsm_35dap_to_44dap = `stsm_35_44`)

summary(X)
names(X)

# Save the data frame:
save(X, file = here::here("DataFusion", "FusedData.RData"))
```
